Dissecting the causal question underlying the association between cancer and ADRD: an opportunity to understand sources of bias

### 2021-06-15 04_reports/manuscript.Rmd
latest updates on manuscript

---
title: "Dissecting the causal question underlying the association between cancer and ADRD: an opportunity to understand sources of bias"
author: ""
date: ""
output: [pdf_document]
always_allow_html: true
editor_options: 
  chunk_output_type: console
bibliography: cancer_dementia.bib
---
```{r, echo =  FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(survival)
library(tidyverse)
library(tableone)
library(magrittr)
library(flextable)
source(here::here("02_R", "03_descriptives.R"))
source(here::here("02_R", "table_2.R")) 
load(here::here("03_figs", "plots.RData")) 
source(here::here("02_R", "05_plots.R"))
```

@@

To explore how we might learn about this effect using real-world data on cancer and ADRD, we progressively build a causal directed acyclic graph to connect this particular causal question to the observable data and the assumptions we rely on to study the effect. We exemplify different scenarios with data collected from the Rotterdam Study, a population-based cohort study. We describe the challenges and how they translate into the analytic decisions. Last, we discuss how information on mortality and cause of death can provide insight about the direction of some sources of bias.

## 2. Overview of the causal structure

@@

A challenge that arises by defining cancer diagnosis as the proxy for Pin1 is defining the time zero, the time where follow-up and screening for ADRD would begin after having the exposure measured. The eligibility criteria to join the study may not align with the time cancer diagnosis is measured. This situation may introduce inmortal-time bias [@hernanInmortal2016]. For example, a study performed using data from the Framingham study [@driver2012] defined the exposed group with cancer as those participants with prevalent or incident cancer diagnosis (alternatively defined as "ever cancer" [@hanson2016]). This meant that a participant who had cancer diagnosis over follow-up contributed all their person-time to the cancer arm, including the time prior to the cancer diagnosis. By defining the exposure after the time of inclusion to the cohort, only participants who remain alive and free of ADRD diagnosis over follow-up can be defined as "ever cancer" [@anderson1983, @hernanInmortal2016].

@@

We now move to the challenge of death as a competing event for ADRD, represented in Figure 4. For interpretability we exclude the time-varying process of cancer diagnosis and focus on Pin1 (and cancer diagnosis) as it had been measured in all participants at $t$. Given that some participants may die prior to having a cancer diagnosis over follow-up, we can only have ADRD over follow-up in the individuals who survive long enough to have an ADRD diagnosis. In the causal diagram of Figure 4 we include a node for death after the exposure $P^*t$ has been measured, represented as $D_{t+1}$, this acts as a competing event of $Y_{t+1}$ because if a participant dies by $t + 1$, the participant cannot subsequently develop dementia. Furthermore, since $D_{t+1}$ and $Y_{t+1}$ are events related to aging and its consequences, $C$ represents the shared causes of both events such as cardiovascular conditions. We also include an arrow between $L$ and $D_{t+1}$ following the argument discussed previously.

In the setting where $P_t$ represented the targeted-drug for Pin1, and if this drug had no systemic beneficial or harmful side-effects such as that there is no arrow between $P$ and $D_{t+1}$, a total effect would quantify the effect of $P_t$ on $Y_{t+1}$ that does not include any pathway mediated through $D_{t+1}$[@young2020]. However, in the context of cancer diagnosis as the proxy for Pin1 over-expression, we cannot rule out the effect of cancer diagnosis on death, represented as the arrow between $P^*_t$ and  $D_{t+1}$. In this setting, a total effect of $P_t$ in $Y_{t+1}$ would include the causal pathway mediated by the effect of cancer diagnosis in mortality, which may translate into an inverse association[@young2020].

To isolate the direct effect of $P_t$ in $Y_{t+1}$ through measurement of $P^*_t$ we have at least two alternatives of causal questions we can ask. We can either imagine a causal question where we decompose the effect of cancer by the different mechanisms that affect dementia and death separately[@stensrud2020]. Alternatively, we can define an scenario where death could have been prevented. The latter is defined as the controlled direct effect, where death is considered as a censoring event and it relies on the assumption that we have measured all $C$ to block the pathway $Y_{t+1} \leftarrow C  \rightarrow D_{t+1} \rightarrow P^*_t \rightarrow P_t$. Previous studies have defined death as a censoring event [], although failed to explicitly define how to comply with the independent censoring assumption. Moreover, adjusting for confounders for dementia (such as only adjusting for $L$) may represent partially the adjustment for the competing event of death [@geskus], to block the pathway mediated by $C$ we need to consider covariates that change over time and that affect both nodes, such as cardiometabolic conditions.

@@

We now turn to an application where we show how these challenges translate in analytic decisions, and we will show ways to mitigate or better understand to the best of the available data's abilities. Nevertheless, we acknowledge that cancer diagnosis represents a complex and heterogenous health condition that exceeds the representation of Pin1 expression, thus, at best we may test the sharp null hypothesis rather than estimating the causal effect of Pin1. In this section we conduct an analysis of the cancer-ADRD association that is structured to the best of the available data's abilities to inform the possible effect of Pin1 on all cause-dementia.

## 3. Application to the Rotterdam Study

We use data collected in the Rotterdam Study, a population-based prospective cohort study among persons living in the Ommoord district in Rotterdam, the Netherlands. Recruitment and initial assessments were held between 1990 and 1993; it was later extended between 2000 and 2001 consisting of individuals who had reached the age of 55 years or who had moved into the study area. Participants from first subcohort had follow-up visits between 1993-1995, 1997-1999, 2002-2005, and 2008-2010, second subcohort had follow-up visits between 2004 and 2005, and between 2011 and 2012[@ikram2020]. All participants had data on incident cancer diagnosis and incident dementia diagnosis through follow-up, collected from medical records of general practitioners (including hospital discharge letters) and through linkage with national registries. Date and cause of death was collected on a weekly basis via municipal population registries. Data on clinical outcomes was available until 2015.

We considered as inclusion criteria being between 60 and 70 years old at study entry, without history of cancer diagnosis, free of cognitive decline or with previous history of dementia. Out of 10998 persons who participated at study entry, `r total_n` were considered eligible. Time to cancer diagnosis, time to dementia diagnosis and death status was measured for all participants. All participants were followed from study entry until dementia diagnosis, death or 20 years after their individual baseline date, whichever occurred first.

### 3.1. Methods

We illustrate the association between cancer and dementia diagnosis under two scenarios. Scenario A replicates the setting that defines cancer proxy as _"cancer ever vs. never"_ [@driver2012]. Scenario B defines cancer diagnosis is a time-varying exposure. To address confounding we fit inverse probability treatment weights, stabilized and truncated at 99th percentile. In scenario A, weights were defined as the inverse of the probability of cancer diagnosis conditional on baseline covariates such as age at study entry, sex, educational attainment, cohort, smoking status. In contrast, for scenario B weights were defined to represent the product of the inverse probability of being diagnosed with cancer over time, conditional on time-varying covariate history. This recreates a pseudopopulation in which time-varying covariates do not predict the risk of cancer diagnosis at time _t_ [@hernan2000]. Time varying covariates included age at study entry, sex, ApoE4 status, educational attainment and the time-varying covariates such as smoking status, systolic blood pressure, BMI  and prevalent and incident comorbidities such as: hypertension and diabetes. Further details on modeling specifications and weights assessment are presented as **Supplementary material x**.

Inverse probability censoring weights for death were calculated to relax the independent censoring assumption. In scenario A, weights represent the inverse of the probability of not dying conditional on baseline covariates such as age, educational attainment, ApoE4, and baseline measurements of smoking status, hypertension status, systolic blood pressure, BMI, history of diabetes and cohort. For individuals who died, their censoring weight is zero [@whatif2020]. In scenario B time-varying weights represent the product of the inverse probability of surviving in each year prior to _t_, conditional on the measured shared causes of death and dementia. For an individual who has died by time _t_, the year _t_ censoring weight is zero. Weights were fitted including the same covariates used to estimate weights for time-varying cancer diagnosis, though we additionally added time-varying cancer, stroke, and heart disease diagnosis as predictors for death.

@@

For illustrative and comparative purposes we calculated hazard ratios (HR). Hazards, unlike risks, inherently condition on surviving both dementia and death, as such they will not have a causal interpretation in this case [@young2020]. We additionally calculated the controlled direct effect considering death as an unconditional independent censoring event (as if there were no arrows from $C$ to $D_{t+1}$ and $Y_{t+1}$).

@@

### 3.2. Results

@@

In contrast, had we defined cancer diagnosis as a _time-varying_ proxy for Pin1 over-expression and had we prevented death conditional on covariates, the risk of dementia had participants had a cancer diagnosis over time is higher [RR: `r complete_table[6,4]`; HR: `r complete_table[6,5]`], though confidence intervals cross the null.

>>>>

---
title: "Dissecting the causal question underlying the association between cancer and ADRD: an opportunity to understand sources of bias"
author: ""
date: ""
output: [word_document]
always_allow_html: true
editor_options: 
  chunk_output_type: console
bibliography: cancer_dementia.bib
csl: AnD_journal.csl
---
```{r, echo =  FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(survival)
library(tidyverse)
library(tableone)
library(magrittr)
library(flextable)
source(here::here("02_R", "03_descriptives.R"))
source(here::here("02_R", "table_2.R")) 
load(here::here("03_figs", "plots.RData")) 
source(here::here("02_R", "05_plots.R"))
```

@@

To explore how we might learn about this effect using real-world data on cancer and ADRD, we progressively build a causal directed acyclic graph to connect this particular causal question to the observable data and the assumptions we rely on to study the effect. We exemplify different scenarios with data collected from the Rotterdam Study, a population-based cohort study. We describe the challenges and how they translate into the analytic decisions.

## 2. Methods

### 2.1. Overview of the causal structure

@@

A challenge that arises by defining cancer diagnosis as the proxy for Pin1 is defining the time zero, the time where eligibility criteria is met, where exposure is measured and the time where the screening for ADRD would begin after having the exposure measured. The eligibility criteria to join the study may not align with the time cancer diagnosis is measured. This situation may introduce inmortal-time bias [@hernanInmortal2016]. For example, a study performed using data from the Framingham study [@driver2012] defined the exposed group with cancer as those participants with prevalent or incident cancer diagnosis (alternatively defined as "ever cancer" [@hanson2016]). This meant that a participant who had cancer diagnosis over follow-up contributed all their person-time to the cancer arm, including the time prior to the cancer diagnosis. By defining the exposure after the time of inclusion to the cohort, only participants who remain alive and free of ADRD diagnosis over follow-up can be defined as "ever cancer" [@anderson1983, @hernanInmortal2016].

@@

We now move to the challenge of death as a competing event for ADRD, represented in Figure 4. For interpretability we exclude the time-varying process of cancer diagnosis and focus on Pin1 (and cancer diagnosis) as it had been measured in all participants at $t$. Given that some participants may die prior to ADRD diagnosis, we can only measure ADRD over follow-up in the individuals who survive long enough to have an ADRD diagnosis. In the causal diagram of Figure 4 we include a node for death after the exposure $P^*t$ has been measured, represented as $D_{t+1}$, this acts as a competing event of $Y_{t+1}$ because if a participant dies by $t + 1$, the participant cannot subsequently develop dementia. Furthermore, since $D_{t+1}$ and $Y_{t+1}$ are events related to aging and its consequences, $C$ represents the shared causes of both events such as cardiovascular conditions. We also include an arrow between $L$ and $D_{t+1}$ following the argument discussed previously.

In the setting where $P_t$ represented the targeted-drug for Pin1, and if this drug had no systemic beneficial or harmful side-effects such as that there is no arrow between $P_t$ and $D_{t+1}$, a total effect would quantify the effect of $P_t$ on $Y_{t+1}$ that does not include any pathway mediated through $D_{t+1}$[@young2020]. However, in the context of cancer diagnosis as the proxy for Pin1 over-expression, we cannot rule out the effect of cancer diagnosis on death, represented as the arrow between $P^*_t$ and  $D_{t+1}$. In this setting, a total effect of $P_t$ in $Y_{t+1}$ would include the causal pathway mediated by the effect of cancer diagnosis in mortality, which may translate into an inverse association[@young2020].

To isolate the direct effect of $P_t$ in $Y_{t+1}$ through measurement of $P^*_t$ we have at least two alternatives of causal questions we can ask. We can either imagine a causal question where we decompose the effect of cancer by the different mechanisms that affect dementia and death separately[@stensrud2020]. Alternatively, we can define an scenario where death could have been prevented. The latter is defined as the controlled direct effect, where death is considered as a censoring event and it relies on the assumption that we have measured all $C$ to block the pathway $Y_{t+1} \leftarrow C  \rightarrow D_{t+1} \rightarrow P^*_t \rightarrow P_t$. Previous studies have defined death as a censoring event [], although failed to explicitly define how to comply with the independent censoring assumption. Moreover, adjusting for confounders for dementia (such as only adjusting for $L$) may represent partially the adjustment for the competing event of death [@geskus2016], though to block the pathway mediated by $C$ we need to consider covariates that change over time and that affect both nodes, such as cardiometabolic conditions.

@@

We now turn to an application where we show how these challenges translate in analytic decisions. We will show ways to mitigate or better understand the best of the available data's abilities to inform the possible effect of Pin1 on all cause-dementia.

## 2.2. The Rotterdam Study

We use data collected in the Rotterdam Study, a population-based prospective cohort study among persons living in the Ommoord district in Rotterdam, the Netherlands. Recruitment and initial assessments were held between 1990 and 1993, a second waive of recruitment was held between 2000 and 2001. Participants from first subcohort had follow-up visits between 1993-1995, 1997-1999, 2002-2005, and 2008-2010, second subcohort had follow-up visits between 2004 and 2005, and between 2011 and 2012[@ikram2020]. All participants had data on history of cancer and incident dementia diagnosis through follow-up, collected from medical records of general practitioners (including hospital discharge letters) and through linkage with national registries. Date and cause of death was collected on a weekly basis via municipal population registries. Data on clinical outcomes was available until 2015.

We defined as inclusion criteria being between 60 and 70 years old at study entry, without history of cancer diagnosis, free of cognitive decline or with previous history of dementia. Out of 10998 persons who participated at study entry, `r total_n` were considered eligible. Time to cancer diagnosis, time to dementia diagnosis and death status was measured for all participants. All participants were followed from study entry until dementia diagnosis, death or 20 years after their individual baseline date, whichever occurred first.

### 2.3. Statistical Methods

We illustrate the association between cancer and dementia diagnosis under two scenarios. Scenario A replicates the setting that defines cancer proxy as _"cancer ever vs. never"_ [@driver2012]. Scenario B defines cancer diagnosis is a time-varying exposure. To address confounding we fit inverse probability treatment weights, stabilized and truncated at 99th percentile. In scenario A, weights were defined as the inverse of the probability of cancer diagnosis conditional on baseline covariates such as age at study entry, sex, educational attainment, cohort, smoking status. In contrast, for scenario B weights were defined to represent the product of the inverse probability of being diagnosed with cancer over time, conditional on the time-varying covariate history prior to cancer diagnosis. This recreates a pseudopopulation in which time-varying covariates do not predict the risk of cancer diagnosis at time _t_ [@hernan2000]. Time varying covariates included age at study entry, sex, ApoE4 status, educational attainment and the time-varying covariates such as smoking status, systolic blood pressure, BMI  and prevalent and incident comorbidities such as: hypertension and diabetes. Further details on modeling specifications and weights assessment are presented as **Supplementary material x**.

Inverse probability censoring weights for death were calculated to relax the independent censoring assumption. In scenario A, weights represent the inverse of the probability of not dying conditional on cancer diagnosis (ever vs. never), baseline covariates such as age, educational attainment, ApoE4, and baseline measurements of smoking status, hypertension status, systolic blood pressure, BMI, history of diabetes and cohort. For individuals who died, their censoring weight is zero [@whatif2020]. In scenario B time-varying weights represent the product of the inverse probability of surviving in each year prior to _t_, conditional on the measured shared causes of death and dementia. For an individual who has died by time _t_, the year _t_ censoring weight is zero. Weights were fitted including the same covariates used to estimate weights for time-varying cancer diagnosis, though we additionally added time-varying cancer, stroke, and heart disease diagnosis as predictors for death.

@@

For illustrative and comparative purposes we calculated hazard ratios (HR). Hazards, unlike risks, inherently condition on surviving both dementia and death, as such they will not have a causal interpretation in this setting [@young2020]. We additionally calculated the controlled direct effect considering death as an unconditional independent censoring event (as if there were no arrows from $C$ to $D_{t+1}$ and $Y_{t+1}$).

@@

## 3. Results

@@

In contrast, had we defined cancer diagnosis as a _time-varying_ proxy for Pin1 over-expression and had we prevented death conditional on covariates, the risk of dementia had participants had a cancer diagnosis over time is higher [RR: `r complete_table[6,4]`; HR: `r complete_table[6,5]`], though confidence intervals cross the null. In the extreme scenarios of dependence between death and dementia over follow-up, bounds are wide and include the null value ()

## 4. Discussion

Several observational studies have defined “cancer diagnosis” as exposure, although this does not represent a target for potential intervention or a modifiable risk factor. Instead, this variable is commonly used to represent a mechanism of interest that could not be measured. In this study we define cancer diagnosis as the proxy for Pin1 expression, a potential therapeutic drug-target. By explicitly including Pin1 as part of the research question we connect the unmeasured mechanism of interest to the observed data outlining the data generating mechanisms. This process allows to identify and disentangle potential sources of bias and can guide analytic decisions. We show how results of the association between cancer diagnosis and dementia can change substantially according to alternative, yet explicit, assumptions.

For instance, a key challenge of cancer diagnosis as a proxy for Pin1 is the incapacity of defining a clear time zero[@hernanInmortal2016]. In the setting where cancer diagnosis is defined as "ever vs. never", this definition implicitly introduces immortal time bias, only participants who survived and remain free of dementia over follow-up can be considered for the cancer arm. All results pertaining to this definition showed a substantial inverse association between cancer diagnosis and dementia, in contrast to cancer diagnosis as a time-varying process which shifted point estimates close to the null. In line with the latter definition, including time-varying covariates is possible with well-known statistical methods such as inverse probability weighting, as opposed to previous studies that only considered baseline information. Although we attempt to prevent this bias with statistical methods, we can only fully prevent it by having a clear definition of time-zero. This definition does not depend on the collected data nor in analytic decisions. It relies on a deeper discussion related to when would be the best moment to measure this variable and to what purpose. Thus, we hope that these unsolved questions guide future discussions and data collection efforts.

In contrast, death as competing event is a challenge that has a straightforward solution, which begins by choosing the causal parameters (or estimands) of interest [@young2020, @rojas_medrxiv]. In this study we chose the controlled direct effect, which represents the effect of Pin1 (or cancer) in a setting where death could have been prevented. To address this question, death is defined as a censoring event and relies on the independent censoring assumption. As opposed to prior studies that treated censoring as ignorable[@frain2017], we show how point estimates change substantially when we include IPCW for death to relax this assumption. In contrast to prior studies who have defined "survival bias" or "competing event bias" and have proposed different estimators to account for them[@ospina2020], we conceptualize bias as not meeting the independent censoring assumption conditional on covariates[@vangeloven2014, @willems2018, @young2020]. Bounds to assess extreme scenarios of dependence between death and dementia[@peterson1976] illustrate the wide range of possible point estimates that cross the null. This shows that even with the effort of adjusting for time-varying covariates, we may be far from meeting this assumption and thus better efforts to measure shared causes of dementia and death are needed.

Pin1 is only one of the several mechanisms proposed about the inverse cancer-dementia relationship. However, each mechanism and its connection to collected data may have a different causal representation. For example, if the intention was to explain the effect of different chemotherapeutics and how these affect brain, the eligibility criteria would change and the causal graph would represent the downstream of cancer diagnosis. Likewise, an alternative causal question related to death as a competing event is the separable effects, which avoid evoking consideration of implausible scenarios that were death is eliminated[@stensrud2020]. The separable effects would allow the decomposition of Pin1 related mechanisms that affect neurodegeneration and tumor proliferation, that could be confirmed with future studies on modified treatments. Last, this case-study extends to other questions that study the effect of one disease in the risk of another disease to understand the biological mechanisms behind them.

Nevertheless, we acknowledge that cancer diagnosis represents a complex and heterogeneous health condition that exceeds the representation of Pin1 expression, thus, at best we may test the sharp null hypothesis rather than estimating the causal effect of Pin1.
